---
title: nihexporter README
author: Jay Hesselberth <jay.hesselberth@gmail.com>
output: md_document
---

# Overview

The `nihexporter` package provides a minimal set of data from NIH EXPORTER
<http://exporter.nih.gov/default.aspx>, which contains information on NIH
biomedical research funding from 1985-2014 (and continues monthly in a given fiscal year).

Information about specific columns in the tables is at <http://exporter.nih.gov/about.aspx>.

The package contains the following tables:

* `projects`: provides data on funded projects by NIH.

* `project.pis`: links project numbers (`project.num`) to PI ID (`pi.id`), which can used in NIH REPORTER searches

* `project.orgs`: links DUNS numbers (`org.duns`) from `projects` table to information on specific organizations 

* `publinks`: links Pubmed IDs (`pmid`) to project numbers (`project.num`)

There are also a few helper variables that make exploratory analysis a bit easier:

* `nih.institutes`: 27 NIH institutes in two-letter format 

## Data summary

There is a lot of data in NIH EXPORTER, so this package aims to provide a
minimal set of data without being too unwieldy. There are download and
import scripts in the `data-raw/` directory in the
package.

Because `total.cost` is only available from fiscal year 2000 and onward, only data from those years is provided in the `projects` table. The `publinks` table goes back to 1985.

## Install

Install the `nihexporter` package from github with:
  
```
devtools::install_github("jayhesselberth/nihexporter")
```

# Examples

```{r, load.libs, message=FALSE}
library(dplyr)
library(knitr)
library(ggplot2)
library(nihexporter)
```

List the all-time most expensive grants from each institute:

```{r expensive.projects}
expensive.projects <- projects %>%
  select(project.num, institute, total.cost, project.start, project.end) %>%
  group_by(project.num, institute) %>%
  summarise(overall.cost = sum(total.cost, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(institute) %>%
  arrange(desc(overall.cost)) %>%
  slice(1:1) %>%
  ungroup() %>%
  arrange(desc(overall.cost)) %>%
  mutate(cost.in.billions = overall.cost / 1e9)

head(expensive.projects)
```

Let's look at the amounts spent on R01 grants at each NIH institute. Note this filters for NIH institutes.

```{r grant.costs, message = FALSE, warning=FALSE}
grant.costs <- projects %>% 
  filter(institute %in% nih.institutes) %>%
  filter(activity == 'R01' & total.cost > 0) %>%
  select(institute, total.cost)

library(scales)
grant.costs %>%
  ggplot(aes(reorder(institute, total.cost, median, order=TRUE), total.cost)) +
  geom_boxplot(outlier.shape = NA) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 8e5), labels = comma) +
  ylab('Total cost (dollars)') +
  xlab('NIH institute') + 
  ggtitle('Total cost of R01 grants from 2000-2014')
```

See the [vignette](http://rpubs.com/jayhesselberth/nihexporter-vignette) or the [vignette source](vignettes/nihexporter.Rmd) for more examples.
