---
title: '`nihexporter`: an R package for analyzing NIH funding data'
author: "Jay R. Hesselberth"
date: "March 14, 2015"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    theme: cerulean
---

```{r, echo=FALSE}
  library(knitcitations); cleanbib()
  cite_options(citation_format = "pandoc", check.entries=FALSE)
  library(bibtex)
```

# Abstract

## Motivation

The analysis of NIH funding trends is important to illustrate how
biomedical research funding in the United States. However, these analyses
are conducted ad hoc by the institutes themselves. The NIH provides free
access to NIH funding data via NIH EXPORTER. 

## Results

We developed the `nihexporter` R package, which provides access to NIH
EXPORTER data. We developed several analysis vignettes that show funding
trends across NIH institues over 15 years, and highlight important
differences in how institutes change their funding profiles. The package illustrates exploratory data analysis with the `dplyr` and `ggplot2` R packages Individuals and institutions with their own NIH funding can perform self-study and
analysis with respect to other grants.

## Availability

The `nihexporter` R package can be installed via
[github](https://github.com/jayhesselberth/nihexporter). The source for
vignettes on institute trends, grant productivity and XXX are available on
rpubs.com. We also provide interactive data analysis of NIH funding data
via an R Shiny app [].

## Implementation

The `nihexporter` package is implemented in the R Statistical Computing
Environment.

## Contact

Jay Hesselberth <jay.hesselberth@gmail.com>

## Supplementary Information

  * Vignettes - 
  
  * Interactive analysis - 

# Introduction

# Methods

# Results and Discussion

## Funding trends

## Grant producitivity

# References

```{r, warning=FALSE, message=FALSE, echo=FALSE} 
  # write.bibtex(file="references.bib")
```
