---
title: "FundingMaps"
author: "Ashley Pacheco and Mary Sweet"
date: "May 19, 2015"
output: html_document
---

## libraries

```{r}
library(nihexporter)
library(dplyr)
library(maps)
library(ggplot2)
library(scales)
```

## perform two left joins in order to obtain data frame 
## with fiscal.year, org.state, and total.cost

```{r}
#first, left join on org.dun
org.dun_join <- left_join(project_orgs, projects)

#second, left join on project.num
all_join <- left_join(org.dun_join, project_io)
```
## replace state abbreviations will full state name so that maps function can work

```{r}
st_codes <- data.frame(
  org.state=as.factor(c("AK", "AL", "AR", "AZ", "CA", "CO", "CT", "DC", "DE", "FL", "GA",
                        "HI", "IA", "ID", "IL", "IN", "KS", "KY", "LA", "MA", "MD", "ME",
                        "MI", "MN", "MO", "MS",  "MT", "NC", "ND", "NE", "NH", "NJ", "NM",
                        "NV", "NY", "OH", "OK", "OR", "PA", "PR", "RI", "SC", "SD", "TN",
                        "TX", "UT", "VA", "VT", "WA", "WI", "WV", "WY")),
  state=as.factor(c("alaska","alabama","arkansas","arizona","california","colorado",
                    "connecticut","district of columbia","delaware","florida","georgia",
                    "hawaii","iowa","idaho","illinois","indiana","kansas","kentucky",
                    "louisiana","massachusetts","maryland","maine","michigan","minnesota",
                    "missouri","mississippi","montana","north carolina","north dakota",
                    "nebraska","new hampshire","new jersey","new mexico","nevada",
                    "new york","ohio","oklahoma","oregon","pennsylvania","puerto rico",
                    "rhode island","south carolina","south dakota","tennessee","texas",
                    "utah","virginia","vermont","washington","wisconsin",
                    "west virginia","wyoming")))
all_join_states <- left_join(st_codes, all_join)
```

## omit all NA values;
## because numbers are too large, divide by a million
## obtain data table with org.state and total cost in millions for 2000-2012 (no 2011 data)
## obtain data table with org.state, fiscal.year, and total cost in millions for that year

```{r}
map_data_all_years <- all_join_states %>%
  na.omit() %>%
  mutate(M.cost = (total.cost/1000000)) %>%
  group_by(state) %>%
  summarise(sum(M.cost))

#rename columns
names(map_data_all_years) <- c("region", "value")

map_data_by_year <- all_join_states %>%
  na.omit() %>%
  mutate(M.cost = (total.cost/1000000)) %>%
  group_by(fiscal.year, state) %>%
  summarise(sum(M.cost))
#rename columns
names(map_data_by_year) <- c("fiscal.year", "region", "value")
```

## plot US map showing total costs for the 12-year range 2000-2012

```{r}
states <- map_data("state")
choro <- merge(states, map_data_all_years, by="region")
choro <- choro[order(choro$order), ]
choro$value <- cut(choro$value, breaks = c(0,100,500,1000,5000,10000,20000,68000))
plot_all_years <- ggplot(choro, aes(x=long, y=lat, group=group)) +
  geom_polygon(aes(fill = value), colour = alpha("white", 1/2), size = 0.2) +
  scale_fill_brewer(palette = "GnBu",
                    name="Funding (in billions)", 
                    labels=c("<1", "1-5", "5-10", "10-50", "50-100", "100-200", ">200")) +
  ggtitle("Total funding 2000-2012") +
  geom_path(data = states, colour = "white", size = .25) +
  theme(axis.text=element_blank(),
        axis.ticks=element_blank(),
        axis.title=element_blank(),
        panel.background=element_blank())
plot_all_years
```


#plot US maps showing total costs per year for 2000-2012

```{r}
states <- map_data("state")
choro <- merge(states, map_data_by_year, by="region")
choro <- choro[order(choro$order), ]
choro$value <- cut(choro$value, breaks = c(0,1,25,50,500,2500,5000,11000))
plot_by_year <- ggplot(choro, aes(x=long, y=lat, group=group)) +
  geom_polygon(aes(fill = value), colour = alpha("white", 1/2), size = 0.2) +
  scale_fill_brewer(palette = "GnBu",
                    name="Funding (in millions)", 
                    labels=c("<1", "1-25", "25-50", "50-500", "500-2500", "2500-5000", ">5000")) +
  ggtitle("Total funding per year") +
  geom_path(data = states, colour = "white", size = .25) +
  theme(axis.text=element_blank(),
        axis.ticks=element_blank(),
        axis.title=element_blank(),
        panel.background=element_blank()) +
  facet_wrap(~fiscal.year)
plot_by_year
```


