---
title: '`nihexporter`: an R package for analyzing NIH funding data'
author: "Jay R. Hesselberth, Erin Baschal, Nick Ellinwood, Ashley Pacheco, Mary Sweet, Yuying Wang"
date: 'Last update: `r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    toc: yes
  word_document: default
---

```{r echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(scales)
library(stringr)
library(dplyr)
library(ggplot2)
library(nihexporter)
library(stringr)
```

# Abstract

## Motivation

The National Institutes of Health (NIH) is the major source of federal funding for biomedical research in the United States. Analysis of past and current NIH funding can illustrate funding trends and identify productive research topics, but these analyses are conducted *ad hoc* by the institutes themselves and only provide a small glimpse of the available data. The NIH provides free access to funding data via NIH EXPORTER, but no tools have been developed to enable analysis of this data.

## Results

We developed the `nihexporter` R package, which provides access to [NIH EXPORTER](http://exporter.nih.gov/) data. We used the package to develop several analysis vignettes that show funding trends across NIH institutes over 15 years and highlight differences in how institutes change their funding profiles. Investigators and institutions can use the package design to perform self-studies of their own NIH funding.

## Availability

The `nihexporter` R package can be installed via [github](https://github.com/jayhesselberth/nihexporter). Exploratory analysis was done using RStudio with the `dplyr`, `ggplot2` and `knitr` packages. A static vignettes on institute trends, grant productivity and XXX are available on rpubs.com. We also provide interactive data analysis of NIH funding data via using R Shiny app [].

## Implementation

The `nihexporter` package is implemented in the R Statistical Computing Environment.

## Contact

Jay Hesselberth <jay.hesselberth@gmail.com>, University of Colorado School of Medicine

## Supplementary Information

  * [Vignettes](http://rpubs.com/jayhesselberth/nihexporter-vignettes)
  
  * [Interactive analysis with Shiny](http://shinyapps.io/jayhesselberth/nihexporter)

# Introduction

The National Institutes of Health (NIH) is the major source federal funds for biomedical research in the United States. The NIH budget is approved by Congress each year, and in fiscal year 2014, the total NIH budget was 30.3 billion dollars (REF). The NIH is divided into 25 institutes, each with its own focus and mission. For example, National Cancer Institue (NCI) focuses on malignant diseases; the National Institute for Allergy and Immune Disease focuses on the immune system and transmissible diseasei; and the National Institute for General Medical Sciences focuses on basic research without a specific disease focus. Each institute negotiates with the NIH director (currently Francis Collins, MD, PhD) for its yearly budget, with budget institutes ranging from millions to several billion dollars.

The NIH provides funds through competetive grants written by internal and external investigators, and the funds associated with these grants can be divided into 'direct' and 'indirect' costs. **Direct costs** are funds that are given to an investigator (or group of investigators) to conduct their proposed research. These funds buy supplies for the experiments and pay the salaries of people to do the work.

By contrast, **indirect costs** are funds that are paid to institutions associated with investigators, and are used to "keep the lights on": they pay for infrastructure costs. However, the "indirect cost recovery" (ICR) rate of each institution, the fraction of each award the institute receives, is congressionally mandated, and there is a wide range in ICR rates. Some of the highest ICR rates are close to 100%, meaning that for every dollar an investigator receives, the institutions receive an equal amount. Moreover, the use of these funds is at the discretion of the institution; a practice that has previously lead to [abuse of these funds](http://news.stanford.edu/pr/94/941018Arc4090.html).

NIH funding is an investment strategy: the institutes invest money in specific research areas, hoping for future returns in the form of publications, patents and skilled trainees. As with any investment strategy, a periodic review can help rebalance the portfolio in order to maximize returns. Analysis of NIH funding data has been performed internally by the NIH, or by contracted third-parties. Several of these analyses have highlighted funding trends and suggested metrics to gauge the 'return' on the NIH 'investment'. For example, "productivity" can be examined as a function of the number of publications produced by grants per dollar of "direct costs".

# Methods

We downloaded NIH funding data from the NIH EXPORTER website in comma-separated value (CSV) format and parsed these data into R data files that each contain specific information:

  * `projects` has information about projects in each fiscal year, keyed by `project.num`
  * `project_orgs` has information about project organizations
  * `project_pis` links PI information to a project
  * `publinks` table links PubMed IDs to `project.num`
  * `publications` links project IDs to PubMed IDs
  * `patents` link patent IDs to project IDs

We also provide several precomputed variables and tables that enable quick and easy exploratory analysis:

  * `nih.institutes`: Two-letter format for 27 NIH institutes
  * `project_io`: This table contains pre-computed values for overall proeject cost (`project.cost`), as well as the number of publications (`n.pubs`) and patents (`n.patents`).
  
We developed several analyses using the `dplyr`, `ggplot2` and `knitr` packages in the RStudio environment. We also developed interactive analyses using the `shiny` package.

# Results and Discussion

## Package data

* Total costs: NIH EXPORTER provides access to the total costs of each grant in each fiscal year; it does not report direct and indirect costs. Indirect costs can be retrieved from other sources [], but are provided by the package only for reference purposes. Calculations that attempt to normalize by indirect costs should be interpreted with appropriate care.

## Examples

Using the `nihexporter` package, one can easily generate summary plots of the total costs of grants at each institute over time.

For example, we can examine the total costs spent on specific grants and specific institutes over time.

```{r example1_institute_funds, message = FALSE, warning = FALSE, echo = FALSE, fig.width = 7, fig.cap = 'Example 1: Funds for specific types of grants at the GM, CA and AI institutes'}
select_activities <- c('R01', 'P01', 'R21', 'P30')
select_inst <- c('GM', 'CA', 'AI')

inst_costs <- projects %>%
  filter(institute %in% select_inst & activity %in% select_activities) %>%
  select(project.num, institute, activity, fiscal.year, total.cost) %>%
  group_by(project.num, activity, institute, fiscal.year) %>%
  summarize(fy.cost = sum(total.cost, na.rm = TRUE))

ggplot(inst_costs, aes(x = factor(fiscal.year),
                       y = fy.cost, 
                       fill = activity)) +
  geom_boxplot(outlier.size = 0.2) + 
  facet_wrap(~ institute) + 
  scale_y_log10(limits = c(9e4, 1e7),
                labels = trans_format("log10", math_format(10^.x))) + 
  annotation_logticks(sides = "l",
                      scaled = TRUE) +
  scale_fill_brewer(palette="Dark2") +
  xlab('') +
  ylab('Total grant dollars, log-scaled') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))
```

## Funding trends

## Grant producitivity

When [Jeremy Berg](http://www.nigms.nih.gov/about/director/pages/berg.aspx) was head of the Institute of General Medical Sciences (NIGMS) from 2003-2011, he routinely provided analysis of funding trends at NIGMS in his "Feedback Loop" blog. One of these measured the productivity per grant dollar by measuring its "scholarly output" (*i.e.*, publications) as a function of direct costs. In this [plot](http://loop.nigms.nih.gov/2010/09/measuring-the-scientific-output-and-impact-of-nigms-grants/) there is an increase in productivity per dollar, until an inflection point at $700K, after which the number of publications *drops*, suggesting a negative influence of grant money on scholarly output. This was interesting and covered [here](http://www.nature.com/news/2010/101116/full/468356a.html).

Here we flesh out this analysis and look at how all institutes perform by this measure (Berg, and now Lorsch, only analyzed GM). One caveat is that we only have access to `total.cost` in NIH EXPORTER, so the numbers include indirect costs. But, this is real cost to the tax-payer.

Here are several addition comparisons, suggested by readers of Berg's original [blog post](http://loop.nigms.nih.gov/2010/09/measuring-the-scientific-output-and-impact-of-nigms-grants/).

### invesigator-intitiated versus RFA-initiated R01

### Comparison of grant programs

The NIH provides funds through differet [grant programs](http://grants.nih.gov/grants/funding/funding_program.htm):

  * `research`: investigator-intitiated, actvities begin with `R` (e.g., R01)
  
  * `program`: activities begin with `P` (e.g. P01)
  
  * `contract`: actvities begin with `U` (e.g. U54)

#### Funding distribution 

```{r example2_activity_comparisons, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = 'Example 2. Comparison of research, program and contract costs over time'}

# from http://grants.nih.gov/grants/funding/funding_program.htm
research_projects <- projects %>% filter(grepl('^R', activity)) %>% select(project.num)
program_projects <- projects %>% filter(grepl('^P', activity)) %>% select(project.num)
contract_projects <- projects %>% filter(grepl('^U', activity)) %>% select(project.num)

grant_costs <- projects %>%
  filter(institute %in% nih.institutes) %>%
  select(project.num, institute, fiscal.year, total.cost)

research_costs <- grant_costs %>%
  semi_join(research_projects) %>%
  group_by(project.num, institute, fiscal.year) %>%
  summarize(grant.cost = sum(total.cost, na.rm = TRUE)) %>%
  mutate(type = 'research')

program_costs <- grant_costs %>%
  semi_join(program_projects) %>%
  group_by(project.num, institute, fiscal.year) %>%
  summarize(grant.cost = sum(total.cost, na.rm = TRUE)) %>%
  mutate(type = 'program')

contract_costs <- grant_costs %>%
  semi_join(contract_projects) %>%
  group_by(project.num, institute, fiscal.year) %>%
  summarize(grant.cost = sum(total.cost, na.rm = TRUE)) %>%
  mutate(type = 'contract')

combined_tbl <- bind_rows(research_costs, program_costs, contract_costs)

ggplot(combined_tbl, aes(x = factor(fiscal.year),
                         y = grant.cost,
                         fill = type)) +
  geom_boxplot(outlier.shape = NA) + 
  scale_y_log10(limits = c(1e4,1e7),
                labels = trans_format("log10", math_format(10^.x))) +
  facet_wrap(~institute) +
  scale_fill_brewer(palette = "Dark2") + 
  xlab('') +
  ylab('Total grant costs, log-scaled') +
  theme(axis.text.x = element_text(angle = 90))

```

#### Productivity

First we calculate the total cost of each R01 grant.

```{r project_productivity, message = FALSE, warning = FALSE}
project_costs <- projects %>%
  filter(activity == 'R01') %>%
  select(project.num, institute) %>%
  left_join(project_io) %>%
  filter(project.cost > 1e4)

project_costs
```

Now we need to measure grant productivity. Cross-referencing these grants with the `project_io` returns the `project.cost`, `n.pubs` and `n.patents`.

```{r plot_dollar_bins, warning = FALSE, message = FALSE, fig.width=7, fig.height=7}
tile.counts <- dollar_grant_productivity %>%
  group_by(institute, dollar.tile) %>%
  summarize(tile.counts = n())

productivity_with_counts <- dollar_grant_productivity %>%
  left_join(tile.counts)

productivity_with_counts %>%
  # need to remove higher tiles because there are too few grants
  filter(dollar.tile <= 13) %>%
  ggplot(aes(x = factor(dollar.tile), y = n.pubs, fill = tile.counts)) +
  geom_boxplot() +
  scale_x_discrete(labels = breaks / 1e6) + 
  theme(axis.text.x = element_text(angle=45, vjust=0.8)) + 
  scale_y_log10() +
  facet_wrap(~ institute) +
  ylab('Number of publications') +
  xlab('Total costs (minimum, in millions)')
```

### Intramural versus extramural grants

The [Center for Scientific Review](http://public.csr.nih.gov/Pages/default.aspx) would need to publicly release (anonymized) scores to enable such an analysis.

### Geographical distribution

Geographical distribution of grant dollar is easily visualized using the package.

```{r geography, warning = FALSE, message = FALSE}
library(maps)
library(leaflet)

state.data <- data.frame(org.state = state.abb,
                         state.name = str_to_lower(state.name))

state_funding <- projects %>%
  select(project.num, org.duns) %>%
  left_join(project_orgs) %>%
  select(project.num, org.state) %>%
  left_join(state.data) %>%
  select(project.num, state.name) %>%
  left_join(project_io)

```

# References
