---
title: "nihexporter"
author: "Jay Hesselberth <jay.hesselberth@gmail.com>"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nihexporter}
  %\VignetteEngine{knitr::rmarkdown} \usepackage[utf8]{inputenc}
---

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` setion of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(nihexporter)
```

```{r}
projects %>%
  group_by(fiscal.year, institute) %>%
  summarize(yearly.cost = sum(total.cost, na.rm = TRUE)) %>%
  filter(yearly.cost > 0) %>%
  arrange(yearly.cost) %>%
  ggplot(aes(x = fiscal.year, y = yearly.cost)) +
  geom_line() +
  facet_wrap(~ institute)
```

```{r biggest.losers}
projects %>%
  filter(total.cost > 0 & activity == 'R01') %>%
  group_by(project.num) %>%
  summarise(overall.cost = sum(total.cost)) %>%
  left_join(publinks) %>%
  filter(!is.na(pmid)) %>%
  group_by(project.num, overall.cost) %>%
  summarize(n.pubs = n()) %>%
  mutate(cost.per.pub = overall.cost / n.pubs) %>%
  ungroup() %>%
  arrange(desc(cost.per.pub)) %>%
  head(50) %>%
  kable()
```

```{r}
projects %>%
  group_by(fiscal.year, institute) %>%
  summarize(yearly.cost = sum(total.cost)) %>%
  head()
```

```{r spending.over.time}
projects %>%
  group_by(institute, activity, fiscal.year) %>%
  summarize(total = sum(total.cost, na.rm = TRUE)) %>%
  filter(total > 0) %>%
  ggplot(aes(x = fiscal.year, y = total, color = institute)) +
  geom_line() +
  facet_wrap(~ activity) +
  scale_y_log10()
```

