---
title: "nihexporter"
author: "Jay Hesselberth <jay.hesselberth@gmail.com>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{nihexporter examples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Overview

The `nihexporter` package provides the following tables of data:

* `projects`: information about funded projects

* `project_pis`: PI information (look these up in NIH REPORTER) for each `project.num`

* `project_orgs`: links DUNS numbers (`org.duns`) from `projects` table to information on specific organizations 

* `publinks`: links grants (`project.num`) to PUBMED IDs (`pmid`).

* `patents`: links patent IDs (`patent.id`) to project IDs (`project.num`)

There are also a few helper variables that make exploratory analysis a bit easier:

* `nih.institutes`: 27 NIH institutes in two-letter format 

* `project_output`: pre-computed values for `n.pubs`, `n.patents` and `project.cost`

The `nihexporter` package is avavilable on [github](https://github.com/jayhesselberth/nihexporter).

# Examples

Many of the examples generate project numbers that you can use to look up project specifics in [NIH REPORTER](http://projectreporter.nih.gov/reporter.cfm). We use the amazing `dplyr`, `ggplot2` and `knitr` packages to manipulate and plot the data.

```{r load_libs, message=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(nihexporter)
```

## Award amounts

The `total.cost` column from the `projects` table has many `NA` values.
You need to use `na.rm` in aggregation functions to ensure robust handling of these
values.

Let's look at spending over time for NIGMS in years that have such data:

```{r inst_cost_over_time, message=FALSE, fig.cap = 'Spending at NIGMS'}
cost.over.time <- projects %>%
  select(institute, total.cost, fiscal.year) %>%
  filter(institute == 'GM') %>%
  group_by(fiscal.year, institute) %>%
  summarize(yearly.cost = sum(total.cost, na.rm = TRUE)) %>%
  filter(yearly.cost > 0)

cost.over.time %>%
  ggplot(aes(x = fiscal.year, y = yearly.cost)) +
  geom_line()
```

Let's look at WHERE the money is going. This example illustrates linking of the `project` and `project.orgs` tables via the `org.duns` column.

```{r money_per_institution, message = FALSE, warning = FALSE}
money.per.institution <- projects %>%
  filter(org.duns != '' & activity == 'R01') %>%
  select(org.duns, fiscal.year, total.cost) %>%
  group_by(org.duns, fiscal.year) %>%
  summarise(total.award.billions = sum(total.cost, na.rm = TRUE) / 1e9) %>%
  ungroup() %>%
  arrange(desc(total.award.billions)) %>%
  left_join(project_orgs) %>%
  select(fiscal.year, org.name, total.award.billions)

money.per.institution %>%
  head(10) %>% kable()
```

And WHO are the big winners? Look these PIs up in NIH REPORTER ...

```{r money_per_pi, message = FALSE}

money.per.pi <- projects %>%
  filter(activity == 'R01') %>%
  select(project.num, total.cost) %>%
  group_by(project.num) %>%
  summarise(total.award.millions = sum(total.cost, na.rm = TRUE) / 1e6) %>%
  ungroup() %>%
  arrange(desc(total.award.millions)) %>%
  left_join(project_pis) %>%
  filter(pi.id != '') %>%
  group_by(pi.id) %>%
  summarise(pi.millionaire = sum(total.award.millions)) %>%
  arrange(desc(pi.millionaire))  

money.per.pi %>%
  head(10) %>% kable()
```

## Productivity

In order to measure the "return"" on the money the NIH invests in the research enterprise,
we can measure scholarly output (i.e., publications) per dollar invested.

Here we identify th highest performing grants outside of the R01 category.
Much has been made of the wasteful spending outside of
investigator-initiated research. Here we can see that this is not always
the case.

```{r highperf_no_r01, message=FALSE, warning = FALSE}
high.perf.not.r01 <- projects %>%
  filter(activity != 'R01') %>%
  group_by(project.num) %>%
  summarise(overall.cost = sum(total.cost, na.rm = TRUE)) %>%
  filter(overall.cost > 1e6) %>%
  left_join(publinks) %>%
  filter(!is.na(pmid)) %>%
  group_by(project.num, overall.cost) %>%
  summarize(n.pubs = n()) %>%
  mutate(cost.per.pub = overall.cost / n.pubs) %>%
  ungroup() %>%
  arrange(cost.per.pub)

high.perf.not.r01 %>%
  head(10) %>%
  kable()
```

The highest performer has `r high.perf.not.r01 %>% head(1) %>% select(n.pubs)` publications at a cost of `r high.perf.not.r01 %>% head(1) %>% select(cost.per.pub)` dollars per publication. Here are the Pubmed IDS of some of those publications:

```{r cheap_pubs, echo=FALSE, message=FALSE, wanrning = FALSE}
high.perf.not.r01 %>%
  head(1) %>%
  select(project.num) %>%
  left_join(publinks) %>%
  head(10) %>%
  kable()
```

Here we quantitate the return on R01 investment and identify the most expensive publications, in grant dollars. One might argue that grants with higher `cost.per.pub` are less good investments.

```{r costly_pubs_r01, message=FALSE}
costly.pubs.r01 <- projects %>%
  filter(activity == 'R01') %>%
  group_by(project.num) %>%
  summarise(overall.cost = sum(total.cost, na.rm = TRUE)) %>%
  left_join(publinks) %>%
  filter(!is.na(pmid)) %>%
  group_by(project.num, overall.cost) %>%
  summarize(n.pubs = n()) %>%
  mutate(cost.per.pub = overall.cost / n.pubs) %>%
  ungroup() %>%
  arrange(desc(cost.per.pub))

costly.pubs.r01 %>%
  head(10) %>%
  kable()
```

## Duration

I am always impressed at how long people keep their grants. Let's identify
the longest running R01 projects. This leverages date manipulation with `lubridate`.

```{r grant_stamina, message=FALSE}
long.grants <- projects %>%
  filter(activity == 'R01') %>%
  select(project.num, project.start, project.end) %>%
  group_by(project.num) %>%
  summarize(longest.run = max(project.end) - min(project.start)) %>%
  arrange(desc(longest.run)) %>%
  mutate(in.years = as.numeric(longest.run) / 365)

long.grants %>%
  head(10) %>%
  kable()
```

### Interpretation

**A `r long.grants %>% head(1) %>% select(in.years) %>% round()` year R01?** Wow.
