---
title: "nihexporter"
output:
  rmarkdown::html_vignette: default
  html_document:
    code_folding: hide
vignette: |
  %\VignetteIndexEntry{Overview}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/README-fig-",
  cache.path = "img/README-cache-"
)
```

```{r gif_animation_hook, echo=FALSE, cache=FALSE}
library(knitr)
library(animation)

# copied from gganimate README
 
ani.options(autobrowse = FALSE, interval = 1)

opts_knit$set(animation.fun = function(x, options, format = "gif") {
  x = c(knitr:::sans_ext(x), knitr:::file_ext(x))
  fig.num = options$fig.num
  format = sub("^[.]", "", format)
  fig.fname = paste0(sub(paste0(fig.num, "$"), "*", x[1]), 
                     ".", x[2])
  mov.fname = paste0(sub(paste0(fig.num, "$"), "", x[1]), ".", 
                     format)

  # order correctly
  figs <- Sys.glob(fig.fname)
  figs <- figs[order(as.numeric(stringr::str_match(figs, paste0("(\\d+)\\.", x[2]))[, 2]))]

  animation::im.convert(figs, output = mov.fname)
  
  sprintf("![%s](%s)", options$fig.cap, paste0(opts_knit$get("base.url"), mov.fname))
})

opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE, fig.show = "animate")
```

# Package overview

## Tables

The `nihexporter` package contains several relational tables that contain a variety of information about NIH funding.

* `projects` has information about projects in each fiscal year
* `org_info` has information about project organizations
* `project_orgs` links applications to organizations 
* `project_pis` links principal investigators to projects
* `publinks` links PubMed IDs to projects
* `publications` contains [Relative Citation Ratio](https://icite.od.nih.gov/) values and other information.
* `clinical_studies` has information about [clinical trials](http://clinicaltrials.org)
* `patents` links patents to projects
* `project_io` contains pre-computed values for total project cost (`project.cost`), as well as the number of publications (`n.pubs`) and patents (`n.patents`).

## Functions

* `rcr()` uses PubMed IDs to retrieve [RCR](https://icite.od.nih.gov/) values from the iCite database.

## Variables

* `nih.institutes`: Two-letter format for 27 NIH institutes
  
# Examples

We'll use animated charts to illustrate funding trends at the NIH.

## Project number and cost

First we'll take a look at project number and costs for a few project types (R01, R21, etc.) at a few of the bigger institues (GM, CA, AI, etc.). This first animated graph shows how these variables move over ~15 years.

```{r cost, fig.height=4, fig.show='animate', fig.width=7, message=FALSE, fig.cap='Project costs and numbers'}
library(nihexporter)
library(dplyr)
library(ggplot2)
library(gganimate)
library(cowplot)

insts <- c('GM', 'AI', 'CA', 'HL', 'DK', 'NS')
activities <- c('R01', 'R21', 'P01', 'U01')

costs <- projects %>%
  filter(institute %in% insts & activity %in% activities) %>%
  group_by(fiscal.year, institute, activity) %>%
  summarize(n.projects = n(),
            project.costs = sum(fy.cost, na.rm = TRUE) / 1e6)

gp <- ggplot(costs, aes(x = n.projects,
                        y = project.costs,
                        color = institute,
                        frame = fiscal.year)) +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~ activity) +
  scale_color_brewer(palette = 'Dark2') +
  theme_cowplot() + 
  labs(x = 'Project number',
       y = 'Project costs (millions)',
       title = 'Project cost and number in: ')
 
p <- gp + geom_point(size = 3)
 
gganimate(p)
```

**It is also helpful to view these plots as trajectories to see the path of funding over time.** The following graphs indicate that U01 and R21 projects have had significant recent investment from certain institutes, while R01 and P01 projects have remained relatively constant in number and cost.

```{r cost_trajectories, message=FALSE, fig.width=7, fig.height=4, fig.show='animate', cache=FALSE, fig.cap='Cost trajectories'} 
p <- gp + geom_path(aes(cumulative = TRUE,
                        group = institute),
                    arrow = arrow(length = unit(0.1, "npc")))

gganimate(p)
```

## Funding distribution

Let's look at WHERE the money is going. This example illustrates linking of the `project` and `project.orgs` tables via the `org.duns` column.

```{r money_per_institution, message = FALSE, warning = FALSE}
money.per.institution <- projects %>%
  filter(activity == 'R01') %>%
  left_join(project_orgs) %>%
  select(org.duns, fiscal.year, fy.cost) %>%
  group_by(org.duns, fiscal.year) %>%
  summarise(total.award.billions = sum(fy.cost, na.rm = TRUE) / 1e9) %>%
  ungroup() %>%
  arrange(desc(total.award.billions)) %>%
  left_join(org_info) %>%
  select(fiscal.year, org.name, total.award.billions)

money.per.institution %>%
  head(10) %>% knitr::kable()
```

One can also identiy PIs that have accrued large project dollars. 

```{r money_per_pi, message = FALSE}

money.per.pi <- projects %>%
  filter(activity == 'R01') %>%
  select(project.num, fy.cost) %>%
  group_by(project.num) %>%
  summarise(total.award.millions = sum(fy.cost, na.rm = TRUE) / 1e6) %>%
  ungroup() %>%
  arrange(desc(total.award.millions)) %>%
  left_join(project_pis) %>%
  filter(pi.id != '') %>%
  group_by(pi.id) %>%
  summarise(pi.millionaire = sum(total.award.millions)) %>%
  arrange(desc(pi.millionaire))  

money.per.pi %>%
  head(10) %>% knitr::kable()
```

## Productivity

To measure the "return"" on the money the NIH invests in the research enterprise, we can measure scholarly output (i.e., publications) per dollar invested.

Here we identify th highest performing grants outside of the R01 category. Much has been made of the wasteful spending outside of investigator-initiated research. Here we can see that this is not always the case.

```{r highperf_no_r01, message=FALSE, warning = FALSE}
high.perf.not.r01 <- projects %>%
  filter(activity != 'R01') %>%
  group_by(project.num) %>%
  summarise(overall.cost = sum(fy.cost, na.rm = TRUE)) %>%
  filter(overall.cost > 1e6) %>%
  left_join(publinks) %>%
  filter(!is.na(pmid)) %>%
  group_by(project.num, overall.cost) %>%
  summarize(n.pubs = n()) %>%
  mutate(cost.per.pub = overall.cost / n.pubs) %>%
  ungroup() %>%
  arrange(cost.per.pub)

high.perf.not.r01 %>%
  head(10) %>%
  kable()
```

Here we quantitate the return on R01 investment and identify the most expensive publications, in grant dollars. One might argue that grants with higher `cost.per.pub` are less good investments.

```{r costly_pubs_r01, message=FALSE}
costly.pubs.r01 <- projects %>%
  filter(activity == 'R01') %>%
  group_by(project.num) %>%
  summarise(overall.cost = sum(fy.cost, na.rm = TRUE)) %>%
  left_join(publinks) %>%
  filter(!is.na(pmid)) %>%
  group_by(project.num, overall.cost) %>%
  summarize(n.pubs = n()) %>%
  mutate(cost.per.pub = overall.cost / n.pubs) %>%
  ungroup() %>%
  arrange(desc(cost.per.pub))

costly.pubs.r01 %>%
  head(10) %>%
  knitr::kable()
```

## Duration

I am always impressed at how long people keep their grants. Let's identify the longest running R01 projects. 

```{r grant_stamina, message=FALSE}
long.grants <- projects %>%
  filter(activity == 'R01') %>%
  select(project.num, project.start, project.end) %>%
  group_by(project.num) %>%
  summarize(longest.run = max(project.end) - min(project.start)) %>%
  arrange(desc(longest.run)) %>%
  mutate(in.years = as.numeric(longest.run) / 365)

long.grants %>%
  head(10) %>%
  knitr::kable()
```

