---
title: "nihexporter"
author: "Jay Hesselberth <jay.hesselberth@gmail.com>"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: true
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{nihexporter}
  %\VignetteEngine{knitr::rmarkdown} \usepackage[utf8]{inputenc}
---

# Overview

The `nihexporter` package provides the following tables of data:

* `projects`: information about funded projects

* `project.pis`: PI information (look these up in NIH REPORTER) for each `project.num`

* `publinks`: links grants (`project.num`) to PUBMED IDs (`pmid`).

# Examples

Many of the examples generate project numbers that you can use to look up
project specifics in NIH REPORTER
<http://projectreporter.nih.gov/reporter.cfm>

```{r load.libs, message=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(nihexporter)
```

## Cost of awards

The `total.cost` column from the `projects` table has many `NA` values.
You need to use `na.rm` in aggregation functions to ensure robust handling of these
values.

Let's look at spending over time for NIGMS in years that have such data:

```{r inst.cost.over.time, message=FALSE, fig.cap = 'Spending at NIGMS'}
cost.over.time <- projects %>%
  select(institute, total.cost, fiscal.year) %>%
  filter(institute == 'GM') %>%
  group_by(fiscal.year, institute) %>%
  summarize(yearly.cost = sum(total.cost, na.rm = TRUE)) %>%
  filter(yearly.cost > 0)

cost.over.time %>%
  ggplot(aes(x = fiscal.year, y = yearly.cost)) +
  geom_line()
```

Let's look at WHERE the money is going to:

```{r money.per.institution, message = FALSE}
money.per.institution <- projects %>%
  filter(org.duns != '' & activity == 'R01') %>%
  select(org.duns, fiscal.year, total.cost) %>%
  group_by(org.duns, fiscal.year) %>%
  summarise(total.award.billions = sum(total.cost, na.rm = TRUE) / 1e9) %>%
  ungroup() %>%
  arrange(desc(total.award.billions))

money.per.institution %>%
  head(10) %>% kable()
```

Use govbudgets.com to look up the DUNS number <http://www.govbudgets.com/dunsNumber/042250712>.

Mmmm ... cheese steak ...

And WHO are the big winners?

```{r money.per.pi, message = FALSE}

money.per.pi <- projects %>%
  filter(activity == 'R01') %>%
  select(project.num, total.cost) %>%
  group_by(project.num) %>%
  summarise(total.award.millions = sum(total.cost, na.rm = TRUE) / 1e6) %>%
  ungroup() %>%
  arrange(desc(total.award.millions)) %>%
  left_join(project.pis) %>%
  filter(pi.id != '') %>%
  group_by(pi.id) %>%
  summarise(pi.millionaire = sum(total.award.millions)) %>%
  arrange(desc(pi.millionaire))  

money.per.pi %>%
  head(10) %>% kable()
```

Go Blue ...

## Productivity

The NIH puts a premium on *bang-for-your-buck*, or some measurable return
on the money they invest in the research enterprise. A key way to
quantitate this is by measuring scholarly output (i.e., publications) per
dollar invested.

Here we identify to highest performing grants outside of the R01 category.
Much has been made of the wasteful spending outside of
investigator-initiated research. Here we can see that this is not always
the case ...

```{r high.perf.no.r01, message=FALSE}
high.perf.not.r01 <- projects %>%
  filter(activity != 'R01') %>%
  group_by(project.num) %>%
  summarise(overall.cost = sum(total.cost, na.rm = TRUE)) %>%
  filter(overall.cost > 1e6) %>%
  left_join(publinks) %>%
  filter(!is.na(pmid)) %>%
  group_by(project.num, overall.cost) %>%
  summarize(n.pubs = n()) %>%
  mutate(cost.per.pub = overall.cost / n.pubs) %>%
  ungroup() %>%
  arrange(cost.per.pub)

high.perf.not.r01 %>%
  head(10) %>%
  kable()
```

The highest performer has `r high.perf.not.r01 %>% head(1) %>% select(n.pubs)` publications at a cost of `r high.perf.not.r01 %>% head(1) %>% select(cost.per.pub)` dollars per publication. Here are some of those publications:

```{r cheap.pubs, echo=FALSE, message=FALSE}
high.perf.not.r01 %>%
  head(1) %>%
  select(project.num) %>%
  left_join(publinks) %>%
  head(10) %>%
  kable()
```

Here we quantitate the return on R01 investment. One might argue that
grants with higher `cost.per.pub` are less good investments. Or that the
PIs of those grants do a poor job entering linking their grants to
publications in MyNCBI (no small chore).

```{r costly.pubs.r01, message=FALSE}
costly.pubs.r01 <- projects %>%
  filter(activity == 'R01') %>%
  group_by(project.num) %>%
  summarise(overall.cost = sum(total.cost, na.rm = TRUE)) %>%
  left_join(publinks) %>%
  filter(!is.na(pmid)) %>%
  group_by(project.num, overall.cost) %>%
  summarize(n.pubs = n()) %>%
  mutate(cost.per.pub = overall.cost / n.pubs) %>%
  ungroup() %>%
  arrange(desc(cost.per.pub))

costly.pubs.r01 %>%
  head(10) %>%
  kable()
```

## Duration

I am always impressed at how long people keep their grants. Let's identify
the longest running R01 projects ...

```{r grant.stamina, message=FALSE}
long.grants <- projects %>%
  filter(activity == 'R01') %>%
  select(project.num, project.start, project.end) %>%
  group_by(project.num) %>%
  summarize(longest.run = max(project.end) - min(project.start)) %>%
  arrange(desc(longest.run)) %>%
  mutate(in.years = as.numeric(longest.run) / 365)

long.grants %>%
  head(10) %>%
  kable()
```

**A `r long.grants %>% head(1) %>% select(in.years) %>% round()` year R01?** Wow.
